State of the Practice for Medical Imaging Software

### 2022-10-31 StateOfPractice/Papers/MedImageSoft_SOP-arXiv/MedImageSoft_SOP.tex
Edits to SOP for MI paper

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
\usepackage{amsmath,amsthm}

@@

@@

%\pgfplotsset{compat=1.13}

@@

We present the state of the practice for Medical Imaging software. We selected
29 medical imaging projects from 48 candidates, assessed 10 software qualities
(installability, correctness/ verifiability, reliability, robustness, usability,
maintainability, reusability, understandability, visibility/transparency and
reproducibility) by answering 108 questions for each software project, and
interviewed 8 of the 29 development teams. Based on the quantitative data for
the first 9 qualities, we ranked the MI software with the Analytic Hierarchy
Process (AHP). The top three software products were \textit{3D Slicer},
\textit{ImageJ}, and \textit{OHIF Viewer}. By interviewing the developers, we
identified three major pain points: i) lack of resources; ii) difficulty
balancing between compatibility, maintainability, performance, and security;
and, iii) lack of access to real-world datasets for testing. For future MI
software projects, we propose adopting test-driven development, using continuous
integration and continuous delivery (CI/CD), using git and GitHub, maintaining
good documentation, supporting third-party plugins or extensions, considering
web application solutions, and improving collaboration between different MI
software projects. \wss{Update after the paper has been revised.}

@@

We aim to study the state of software development practice for Medical Imaging
(MI) software.  MI tools use images of the interior of the body (from sources
such as Magnetic Resonance Imaging (MRI), Computed Tomography (CT), Positron
Emission Tomography (PET) and Ultrasound) to provide information for diagnostic,
analytic, and medical applications \citep{FDA2021, enwiki:1034887445,
Zhang2008}.  Figure~\ref{Fig_Example} shows an example image of the brain. Given
the importance of MI software and the high number of competing software
projects, we wish to understand the merits and drawbacks of the current
development processes, tools, and methodologies.  We aim to assess through a
software engineering lens the quality of the existing software with the hope of
highlighting standout examples, understanding current pain points and providing
guidelines and recommendations for future development.

@@

Not only do we wish to gain insight into the state of the practice for MI
software, we also wish to understand the development of research software in
general. We wish to understand the impact of the often cited gap, or chasm,
between software engineering and scientific programming \citep{Kelly2007,
Storer2017}. Although scientists spend a substantial proportion of their working
hours on software development \citep{Hannay2009, Prabhu2011}, many developers
learn software engineering skills by themselves or from their peers, instead of
from proper training \citep{Hannay2009}. \citet{Hannay2009} observe that many
scientists showed ignorance and indifference to standard software engineering
concepts. For instance, according to a survey by \citet{Prabhu2011}, more than
half of their 114 subjects did not use a proper debugger when coding.

To gain insights, we devised 10 research questions, which can be applied to MI,
as well as to other domains, of research software \citep{SmithEtAl2021,
SmithAndMichalski2022}.  We designed the questions to learn about the
community's interest in, and experience with, artifacts, tools, principles,
processes, methodologies, and qualities.  When we mention artifacts we mean the
documents, scripts and code that constitutes a software development project.
Example artifacts include requirements, specifications, user manuals, unit
tests, system tests, usability tests, build scripts, API (Application
Programming Interface) documentation, READMEs, license documents, process
documents, and code.  Once we have learned what MI developers do, we then put
this information in context by contrasting MI software against the trends shown
by developers in other research software communities.  Our aim is to collect
enough information to understand the current pain points experienced by the MI
software development community so that we can make some preliminary
recommendations for future improvements.

We based the structure of the paper on the research questions, so for each
research question below we point to the section that contains our answer.  We
start with identifying the relevant examples of MI software:

\begin{enumerate}
	\item[RQ\refstepcounter{rqnum}\therqnum \label{RQ_WhatProjects}:] What MI
	software projects exist, with the constraint that the source code must be
	available for all identified projects? (Section~\ref{ch_results})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_HighestQuality}:] Which
	of the projects identified in \rqref{RQ_WhatProjects} follow current best
	practices, based on evidence found by experimenting with the software and
	searching the artifacts available in each project's repository?
	(Section~\ref{ch_results})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareHQ2Popular}:] How
	similar is the list of top projects identified in \rqref{RQ_HighestQuality}
	to the most popular projects, as viewed by the scientific community?
	(Section~\ref{Sec_VsCommunityRanking})
    \item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareArtifacts}:] How
	do MI projects compare to research software in general with respect to the
	artifacts present in their repositories?
	(Section~\ref{Sec_CompareArtifacts})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareToolsProjMngmnt}:]
	How do MI projects compare to research software in general with respect to
	the use of tools (Section~\ref{Sec_CompareTools}) for:
	\begin{enumerate} 
		\item [\rqref{RQ_CompareToolsProjMngmnt}.a] development; and,
		\item [\rqref{RQ_CompareToolsProjMngmnt}.b] project management?
	\end{enumerate}
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareMethodologies}:]
	How do MI projects compare to research software in general with respect to
	principles, processes, and methodologies used?
	(Section~\ref{Sec_CompareMethodologies})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_PainPoints}:] What are
	the pain points for developers working on MI software projects?
	(Section~\ref{painpoints})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_ComparePainPoints}:] How
	do the pain points of developers from MI compare to the pain points
	for research software in general? (Section~\ref{painpoints})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Concerns}:] For MI
	developers what specific best practices are taken to address the pain points
	and software quality concerns? (Section~\ref{painpoints})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Recommend}:]
	What research software development practice could potentially address the
	pain point concerns identified in \rqref{RQ_PainPoints}).
	(Section~\ref{ch_recommendations})

@@

We designed a general methodology to assess the state of the practice for SC
software \citep{SmithEtAl2021, SmithAndMichalski2022}. Details can be found in
Section~\ref{ch_methods}.  Our methodology has been applied to MI software
\citep{Dong2021} and Lattice Boltzmann Solvers \citep{Michalski2021}.  This
methodology builds off prior work to assess the state of the practice for such
domains as Geographic Information Systems \citep{smith2018state}, Mesh
Generators \citep{smith2016state}, Seismology software
\citep{Smith2018Seismology}, and Statistical software for psychology
\citep{smith2018statistical}.  In keeping with the previous methodology, we have
maintained the constraint that the work load for measuring a given domain should
be feasible for a team as small as one person, and for a short time, ideally
around a person month of effort. We consider a person month as $20$ working
days ($4$ weeks in a month, with $5$ days of work per week) at $8$ person hours
per day, or $20 \times 8 = 160$ person hours.

With our methodology, we first choose an SC domain (in the current case MI) and
identify a list of about 30 software packages. (For measuring MI we used 29
software packages.)  We approximately measure the qualities of each package by
filling in a grading template. Compared with our previous methodology, the new
methodology also includes repository based metrics, such as the number of files,
number of lines of code, percentage of issues that are closed, etc.  With the
quantitative data in the grading template, we rank the software with the
Analytic Hierarchy Process (AHP) (Section~\ref{ch_background} provides details).
After this, as another addition to our previous methodology, we interview some
development teams to further understand the status of their development process.

@@

\item \textbf{Open Source Software (OSS)} For OSS, the source code is openly
accessible. Users have the right to study, change and distribute it under a
license granted by the copyright holder. For many OSS projects, the development
process relies on the collaboration of different contributors worldwide
\citep{Corbly2014}. Accessible source code usually exposes more ``secrets'' of a
software project, such as the underlying logic of software functions, how
developers achieve their works, and the flaws and potential risks in the final
product. Thus, it brings much more convenience to researchers analyzing the
qualities of the project.

@@

\noindent For example, we measured the usability for 3D Slicer and Ginkgo CADx
as $8$ and $7$, respectively; therefore, on the 9-point scale, 3D Slicer compared
to Ginkgo CADx is 2 and Ginkgo CADX versus 3D Slicer is 1/2, as shown in the
sample AHP calculations (Table~\ref{Tbl_SampleAHP}).

@@

Domain experts may be recruited from academia or industry.  The only
requirements are knowledge of the domain and a willingness to be engaged in the
assessment process.  The Domain Expert does not have to be a software developer,
but they should be a user of domain software.  Given that the domain experts are
likely to be busy people, the measurement process cannot put to much of a burden
on their time.  For the current assessment, our Domain Expert (and paper
co-author) is Dr.\ Michael Noseworthy, Professor of Electrical and Computer
Engineering at McMaster University, Co-Director of the McMaster School of
Biomedical Engineering, and Director of Medical Imaging Physics and Engineering
at St.\ Joseph's Healthcare, Hamilton, Ontario, Canada.

In advance of the first meeting with the Domain Expert, the expert is asked to
create a list of top software packages in the domain.  This is done to help
the expert get in the right mind set in advance of the meeting.  Moreover,
by doing the exercise in advance, we avoid the potential pitfall of the expert
approving the discovered list of software without giving it adequate thought.

@@

We initially identified 48 MI candidate software projects from the literature
\citep{Bjorn2017, Bruhschwein2019, Haak2015}, online articles \citep{Emms2019,
Hasan2020, Mu2019}, and forum discussions \citep{Samala2014}.  The full list of
48 packages is available in \citet{Dong2021}.  To reduce the length of the list
to a manageable number (29 in this case, as given in Section~\ref{ch_results}),
we filtered the original list as follows:

@@

\item We focused on the MI software that provides visualization functions, as
described in Section~\ref{sec_scope}. We removed seven packages that were
toolkits or libraries, such as \textit{VTK}, \textit{ITK}, and \textit{dcm4che}.
We removed another three that were for PACS.

@@

We grade the selected software using the measurement template summarized in
\citet{SmithEtAl2021}.  The template provides measures of the qualities listed
in Section~\ref{sec_software_quality}, except for reproducibility, which is
assessed through the developer interviews (Section~\ref{sec_interview_methods}).
For each software package, we fill-in the template questions. To stay within the
target of 160 person hours to measure the domain, we allocated between 1 to 4
hours for each package. Project developers can be contacted for help regarding
installation, if necessary, but a cap of about 2 hours is imposed on the
installation process, to keep the overall measurement time feasible. An excerpt
of the spreadsheet is shown in Figure~\ref{fg_grading_template_example}.  A
column is included for each measured software package.

\begin{figure}[!ht]
\includegraphics[scale=0.67]{figures/template.pdf}
\caption{Grading template example}
\label{fg_grading_template_example}
\end{figure}

The full template consists of 108 questions categorized under 9 qualities.  The
questions were designed to be unambiguous, quantifiable and measurable with
limited time and domain knowledge. The measures are grouped under headings for
each quality, and one for summary information. The summary information (shown in
Figure~\ref{fg_grading_template_example}) is the first section of the template.
This section summarizes general information, such as the software name, purpose,
platform, programming language, publications about the software, the first
release and the most recent change date, website, source code repository of the
product, number of developers, etc.  We follow the definitions given by
\citet{GewaltigAndCannon2012} for the software categories.  Public means
software intended for public use.  Private means software aimed only at a
specific group, while the concept category is used for software written simply
to demonstrate algorithms or concepts. The three categories of development
models are (open source, free-ware and commercial) are discussed in
Section~\ref{sec_software_categories}.  Information in the summary section sets
the context for the project, but it does not directly affect the grading scores.

For measuring each quality, we ask several questions and the typical answers are
among the collection of ``yes'', ``no'', ``n/a'', ``unclear'', a number, a
string, a date, a set of strings, etc. Each quality is assigned an overall
score, between 1 and 10, based on all the previous questions.  Several of the
qualities use the word ``surface''.  This is to highlight that, for these
qualities in particular, the best that we can do is a shallow measure.  For
instance, we are not currently doing any experiments to measure usability.
Instead, we are looking for an indication that usability was considered by the
developers.  We do this by looking for cues in the documentation, like a getting
started manual, a user manual and documentation of expected user
characteristics.  Below is a summary of how we assess adoption of best practices
by measuring each quality.

@@

\item \textbf{Maintainability} We believe that the artifacts of a project,
including source code, documents, and building scripts, significantly influence
its maintainability. Thus we check each project for the presence of such
artifacts as API documentation, bug tracker information, release notes, test
cases, and build scripts. We also check for the use of tools supporting issue
tracking and version control, the percentages of closed issues, and the
proportion of comment lines in the code.

\item \textbf{Reusability} We count the total number of code files for each
project. Projects with a large number of components potentially provide more
choices for reuse. Furthermore, well-modularized code, which tends to have
smaller parts in separate files, is typically easier to reuse. Thus, we assume
that projects with more code files and less Lines of Code (LOC) per file are
more reusable. We also consider projects with API documentation as delivering
better reusability.

\item \textbf{Surface Understandability} Given that time is a constraint, we
cannot look at all code files for each project; therefore, we randomly examine
10 code files for their understandability. We check the code's style within each
file, such as whether the identifiers, parameters, indentation, and formatting
are consistent, whether the constants (other than 0 and 1) are hardcoded, and
whether the code is modularized. We also check the descriptive information for
the code, such as documents mentioning the coding standard, the comments in the
code, and the descriptions or links for details on algorithms in the code.

@@

For projects on GitHub we manually collect additional information, such as the
numbers of stars, forks, people watching this repository, open pull requests,
closed pull requests, and the number of months a repository has been on GitHub.
We need to take care with the project creation date, since a repository can have
a creation date much earlier than the first day on GitHub.  For example, the
developers created the git repository for \textit{3D Slicer} in 2002, but did
not upload a copy of it to GitHub until 2020. Some GitHub data can be found
using its the GitHub Application Program Interface (API) via the following url:
\textit{https://api.github.com/repos/[owner]/[repository]} where [owner] and
[repository] are replaced by the repo specific values. The number of months a
repository has been on GitHub helps us understand the average change of metrics
over time, like the average new stars per month.

@@

As in \citet{SmithEtAl2016}, Virtual machines (VMs) were used to provide an
optimal testing environments for each package. VMs were used because it is
easier to start with a fresh environment without having to worry about existing
libraries and conflicts. Moreover, when the tests are complete the VM can be
deleted, without any impact on the host operating system. The most significant
advantage of using VMs is to level the playing field. Every software install
starts from a clean slate, which removes ``works-on-my-computer'' errors. When
filling in the measurement template spreadsheet, the details for each VM
are noted, including hypervisor and operating system version.

@@

The repository-based measurements summarize the information we can collect from
on-line resources. This information is incomplete because it doesn't generally
capture the development process, the developer pain points, the perceived
threats to software quality, and the developers' strategies to address these
threats.  Therefore, part of our methodology is to also interview developers.

We based our interviews on a list of 20 questions, which can be found in
\citet{SmithEtAl2021}. Some questions are about the background of the software,
the development teams, the interviewees, and how they organize the projects. We
also ask about the developer's understandings of the users. Some questions focus
on the current and past difficulties, and the solutions the team has found, or
will try. We also discuss the importance and current situations of
documentation. A few questions are about specific software qualities, such as
maintainability, understandability, usability, and reproducibility. The
interviews are semi-structured based on the question list; we ask follow-up
questions when necessary.  The interview process presented here was approved by
the McMaster University Research Ethics Board under the application number 
\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/MACREM/Application.pdf}
{MREB\#: 5219}.

We sent interview requests to all 29 projects using contact information from
projects websites, code repository, publications, and from biographic pages at
the teams's institutions.  In the end nine developers from eight of the projects
agreed to participate: \textit{3D Slicer}, \textit{INVESALIUS 3}, \textit{dwv},
\textit{BioImage Suite Web}, \textit{ITK-SNAP}, \textit{MRIcroGL},
\textit{Weasis}, and \textit{OHIF}. We spent about 90 minutes for each
interview. One participant was too busy to have an interview, so they wrote down
their answers. In one case two developers from the same project agreed to be
interviewed. Meetings were held on-line using either Zoom or Teams, which
facilitated recording and automatic transcription of the meetings. The full
interview answers can be found in \citet{Dong2021}.

@@

Table \ref{tab_final_list} shows the 29 software packages that we measured,
along with summary data collected in the year 2020. We arrange the items in
descending order of LOC. We found the initial release dates (Rlsd) for most
projects and marked the two unknown dates with ``?''. We used the date of the
latest change to each code repository to decide the latest update. We found
funding information (Fnd) for only eight projects.  For the number of
contributors (NOC) we considered anyone who made at least one accepted commit as
a contributor. The NOC is not usually the same as the number of long-term
project members, since many projects received change requests and code from the
community.  With respect to the OS, 25 packages work on all three OSs: Windows
(W), macOS (M), and Linux (L). Although the usual approach to cross-platform
compatibility was to work natively on multiple OSes, five projects achieved
platform-independence via web applications. The full measurement data for all
packages is available in \citet{Dong2021-Data}.

>>>>

%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
\usepackage{amsmath,amsthm}

@@

\newcommand{\CC}{C\nolinebreak\hspace{-.05em}\raisebox{.4ex}{\small\bf
+}\nolinebreak\hspace{-.10em}\raisebox{.4ex}{\small\bf +}}

@@

\pgfplotsset{compat=1.13}

@@

We present the state of the practice for Medical Imaging software. We selected
29 medical imaging projects from 48 candidates, assessed 10 software qualities
(installability, correctness/ verifiability, reliability, robustness, usability,
maintainability, reusability, understandability, visibility/transparency and
reproducibility) by answering 108 questions for each software project, and
interviewed 8 of the 29 development teams. Based on the quantitative data for
the first 9 qualities, we ranked the MI software with the Analytic Hierarchy
Process (AHP). The four top ranked software products are: \textit{3D Slicer},
\textit{ImageJ}, \textit{Fiji}, and \textit{OHIF Viewer}. Our ranking is mostly
consistent with the community's ranking, with four of our top five projects also
appearing in the top five of a list ordered by stars-per-year. \textit{DICOM
Viewer} and \textit{dmv} are exceptions; we ranked them lower than their
estimated popularity because of difficulty with their installation.  Comparison
for tool usage. Comparison for process. From interviewing the developers, we
identified five pain points and two qualities of potential concern:: i) lack of
resources; ii) difficulty balancing between compatibility, maintainability,
performance, and security; and, iii) lack of access to real-world datasets for
testing. For future MI software projects, we propose adopting test-driven
development, using continuous integration and continuous delivery (CI/CD), using
git and GitHub, maintaining good documentation, supporting third-party plugins
or extensions, considering web application solutions, and improving
collaboration between different MI software projects. \wss{Update after the
paper has been revised.}

@@

We aim to study the state of software development practice for Medical Imaging
(MI) software.  MI tools use images of the interior of the body (from sources
such as Magnetic Resonance Imaging (MRI), Computed Tomography (CT), Positron
Emission Tomography (PET) and Ultrasound) to provide information for diagnostic,
analytic, and medical applications \citep{FDA2021, enwiki:1034887445,
Zhang2008}.  Figure~\ref{Fig_Example}, which shows an image of the brain,
highlights the importance and value of MI. Through MI medical practitioners and
researchers can noninvasively gain insights into the human body, including
information on injuries and illnesses. Given the importance of MI software and
the high number of competing software projects, we wish to understand the merits
and drawbacks of the current development processes, tools, and methodologies. We
aim to assess through a software engineering lens the quality of the existing
software with the hope of highlighting standout examples, understanding current
pain points and providing guidelines and recommendations for future development.

@@

Not only do we wish to gain insight into the state of the practice for MI
software, we also wish to understand the development of research software in
general. We wish to understand the impact of the often cited gap, or chasm,
between software engineering and research software \citep{Kelly2007,
Storer2017}. Although scientists spend a substantial proportion of their working
hours on software development \citep{Hannay2009, Prabhu2011}, many developers
learn software engineering skills by themselves or from their peers, instead of
from proper training \citep{Hannay2009}. \citet{Hannay2009} observe that many
scientists showed ignorance and indifference to standard software engineering
concepts. For instance, according to a survey by \citet{Prabhu2011}, more than
half of their 114 subjects did not use a proper debugger when coding.

To gain insights, we devised 10 research questions, which can be applied to MI,
as well as to other domains, of research software \citep{SmithEtAl2021,
SmithAndMichalski2022}.  We designed the questions to learn about the
community's interest in, and experience with, software artifacts, tools,
principles, processes, methodologies, and qualities.  When we mention artifacts
we mean the documents, scripts and code that constitutes a software development
project. Example artifacts include requirements, specifications, user manuals,
unit tests, system tests, usability tests, build scripts, API (Application
Programming Interface) documentation, READMEs, license documents, process
documents, and code.  Once we have learned what MI developers do, we then put
this information in context by contrasting MI software against the trends shown
by developers in other research software communities.  Our aim is to collect
enough information to understand the current pain points experienced by the MI
software development community so that we can make some preliminary
recommendations for future improvements.

We based the structure of the paper on the research questions, so for each
research question below we point to the section that contains our answer.  We
start with identifying the relevant examples of MI software for the assessment
exercise:

\begin{enumerate}
	\item[RQ\refstepcounter{rqnum}\therqnum \label{RQ_WhatProjects}:] What MI
	software projects exist, with the constraint that the source code must be
	available for all identified projects? (Section~\ref{ch_results})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_HighestQuality}:] Which
	of the projects identified in \rqref{RQ_WhatProjects} follow current best
	practices, based on evidence found by experimenting with the software and
	searching the artifacts available in each project's repository?
	(Section~\ref{ch_results})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareHQ2Popular}:] How
	similar is the list of top projects identified in \rqref{RQ_HighestQuality}
	to the most popular projects, as viewed by the scientific community?
	(Section~\ref{Sec_VsCommunityRanking})
    \item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareArtifacts}:] How
	do MI projects compare to research software in general with respect to the
	artifacts present in their repositories?
	(Section~\ref{Sec_CompareArtifacts})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareToolsProjMngmnt}:]
	How do MI projects compare to research software in general with respect to
	the use of tools (Section~\ref{Sec_CompareTools}) for:
	\begin{enumerate} 
		\item [\rqref{RQ_CompareToolsProjMngmnt}.a] development; and,
		\item [\rqref{RQ_CompareToolsProjMngmnt}.b] project management?
	\end{enumerate}
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_CompareMethodologies}:]
	How do MI projects compare to research software in general with respect to
	principles, processes, and methodologies used?
	(Section~\ref{Sec_CompareMethodologies})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_PainPoints}:] What are
	the pain points for developers working on MI software projects?
	(Section~\ref{painpoints})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_ComparePainPoints}:] How
	do the pain points of developers from MI compare to the pain points
	for research software in general? (Section~\ref{painpoints})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Concerns}:] For MI
	developers what specific best practices are taken to address the pain points
	and software quality concerns? (Section~\ref{painpoints})
	\item [RQ\refstepcounter{rqnum}\therqnum \label{RQ_Recommend}:]
	What research software development practice could potentially address the
	pain point concerns identified in \rqref{RQ_PainPoints}?
	(Section~\ref{ch_recommendations})

@@

We designed a general methodology to assess the state of the practice for
research software \citep{SmithEtAl2021, SmithAndMichalski2022}. Details can be
found in Section~\ref{ch_methods}.  Our methodology has been applied to MI
software \citep{Dong2021} and Lattice Boltzmann Solvers \citep{SmithEtAl2022,
Michalski2021}.  This methodology builds off prior work to assess the state of
the practice for such domains as Geographic Information Systems
\citep{smith2018state}, Mesh Generators \citep{smith2016state}, Seismology
software \citep{Smith2018Seismology}, and Statistical software for psychology
\citep{smith2018statistical}.  In keeping with the previous methodology, we have
maintained the constraint that the work load for measuring a given domain should
be feasible for a team as small as one person, and for a short time, ideally
around a person month of effort. We consider a person month as $20$ working days
($4$ weeks in a month, with $5$ days of work per week) at $8$ person hours per
day, or $20 \times 8 = 160$ person hours.

With our methodology, we first choose a research software domain (in the current
case MI) and identify a list of about 30 software packages. (For measuring MI we
used 29 software packages.)  We approximately measure the qualities of each
package by filling in a grading template. Compared with our previous
methodology, the new methodology also includes repository based metrics, such as
the number of files, number of lines of code, percentage of issues that are
closed, etc.  With the quantitative data in the grading template, we rank the
software with the Analytic Hierarchy Process (AHP) (Section~\ref{ch_background}
provides details). After this, as another addition to our previous methodology,
we interview some development teams to further understand the status of their
development process.

@@

\item \textbf{Open Source Software (OSS)} For OSS, the source code is openly
accessible. Users have the right to study, change and distribute it under a
license granted by the copyright holder. For many OSS projects, the development
process relies on the collaboration of different contributors worldwide
\citep{Corbly2014}. Accessible source code usually exposes more ``secrets'' of a
software project, such as the underlying logic of software functions, how
developers achieve their works, and the flaws and potential risks in the final
product. Thus, OSS is suitable for researchers analyzing the qualities of a
project.

@@

\noindent For example, we measured the usability for 3D Slicer and Ginkgo CADx
as $8$ and $7$, respectively; therefore, on the 9-point scale, 3D Slicer compared
to Ginkgo CADx is 2 and Ginkgo CADx versus 3D Slicer is 1/2, as shown in the
sample AHP calculations (Table~\ref{Tbl_SampleAHP}).

@@

Domain experts may be recruited from academia or industry.  The only
requirements are knowledge of the domain and a willingness to be engaged in the
assessment process.  The Domain Expert does not have to be a software developer,
but they should be a user of domain software.  Given that the domain experts are
likely to be busy people, the measurement process cannot put too much of a burden
on their time.  For the current assessment, our Domain Expert (and paper
co-author) is Dr.\ Michael Noseworthy, Professor of Electrical and Computer
Engineering at McMaster University, Co-Director of the McMaster School of
Biomedical Engineering, and Director of Medical Imaging Physics and Engineering
at St.\ Joseph's Healthcare, Hamilton, Ontario, Canada.

In advance of the first meeting with the Domain Expert, they are asked to
create a list of top software packages in the domain.  This is done to help
the expert get in the right mind set in advance of the meeting.  Moreover,
by doing the exercise in advance, we avoid the potential pitfall of the expert
approving the discovered list of software without giving it adequate thought.

@@

We initially identified 48 MI candidate software projects from the literature
\citep{Bjorn2017, Bruhschwein2019, Haak2015}, on-line articles \citep{Emms2019,
Hasan2020, Mu2019}, and forum discussions \citep{Samala2014}.  The full list of
48 packages is available in \citet{Dong2021}.  To reduce the length of the list
to a manageable number (29 in this case, as given in Section~\ref{ch_results}),
we filtered the original list as follows:

@@

\item We focused on the MI software that provides visualization functions, as
described in Section~\ref{sec_scope}. Furthermore, we removed seven packages that were
toolkits or libraries, such as \textit{VTK}, \textit{ITK}, and \textit{dcm4che}.
We removed another three that were for PACS.

@@

We grade the selected software using the measurement template summarized in
\citet{SmithEtAl2021}.  The template provides measures of the qualities listed
in Section~\ref{sec_software_quality}, except for reproducibility, which is
assessed through the developer interviews (Section~\ref{sec_interview_methods}).
For each software package, we fill in the template questions. To stay within the
target of 160 person hours to measure the domain, we allocated between one and
four hours for each package. Project developers can be contacted for help
regarding installation, if necessary, but we impose a cap of about two hours on
the installation process, to keep the overall measurement time feasible.
Figure~\ref{fg_grading_template_example} shows an excerpt of the spreadsheet.
The spreadsheet includes a column for each measured software package.

\begin{figure}[!ht]
\includegraphics[scale=0.66]{figures/template.pdf}
\caption{Grading template example}
\label{fg_grading_template_example}
\end{figure}

The full template consists of 108 questions categorized under 9 qualities.  We
designed the questions to be unambiguous, quantifiable, and measurable with
limited time and domain knowledge. We group the measures under headings for each
quality, and one for summary information. The summary information (shown in
Figure~\ref{fg_grading_template_example}) is the first section of the template.
This section summarizes general information, such as the software name, purpose,
platform, programming language, publications about the software, the first
release and the most recent change date, website, source code repository of the
product, number of developers, etc.  We follow the definitions given by
\citet{GewaltigAndCannon2012} for the software categories.  Public means
software intended for public use.  Private means software aimed only at a
specific group, while the concept category is for software written simply to
demonstrate algorithms or concepts. The three categories of development models
are (open source, free-ware and commercial) are discussed in
Section~\ref{sec_software_categories}.  Information in the summary section sets
the context for the project, but it does not directly affect the grading scores.

For measuring each quality, we ask several questions and the typical answers are
among the collection of ``yes'', ``no'', ``n/a'', ``unclear'', a number, a
string, a date, a set of strings, etc. The grader assigns each quality an
overall score, between 1 and 10, based on all the previous questions.  Several
of the qualities use the word ``surface''.  This is to highlight that, for these
qualities in particular, the best that we can do is a shallow measure.  For
instance, we are not currently doing any experiments to measure usability.
Instead, we are looking for an indication that the developers considered
usability.  We do this by looking for cues in the documentation, like a getting
started manual, a user manual and a statement of expected user characteristics.
Below is a summary of how we assess adoption of best practices by measuring each
quality.

@@

\item \textbf{Maintainability} We believe that the artifacts of a project,
including source code, documents, and building scripts, significantly influence
its maintainability. Thus, we check each project for the presence of such
artifacts as API documentation, bug tracker information, release notes, test
cases, and build scripts. We also check for the use of tools supporting issue
tracking and version control, the percentages of closed issues, and the
proportion of comment lines in the code.

\item \textbf{Reusability} We count the total number of code files for each
project. Projects with numerous components potentially provide more choices for
reuse. Furthermore, well-modularized code, which tends to have smaller parts in
separate files, is typically easier to reuse. Thus, we assume that projects with
more code files and fewer Lines of Code (LOC) per file are more reusable. We also
consider projects with API documentation as delivering better reusability.

\item \textbf{Surface Understandability} Given that time is a constraint, we
cannot look at all code files for each project; therefore, we randomly examine
10 code files for their understandability. We check the code's style within each
file, such as whether the identifiers, parameters, indentation, and formatting
are consistent, whether the constants (other than 0 and 1) are not hardcoded, and
whether the code is modularized. We also check the descriptive information for
the code, such as documents mentioning the coding standard, the comments in the
code, and the descriptions or links for details on algorithms in the code.

@@

For projects on GitHub we manually collect additional information, such as the
numbers of stars, forks, people watching this repository, open pull requests,
closed pull requests, and the number of months a repository has been on GitHub.
We need to take care with the project creation date, since a repository can have
a creation date much earlier than the first day on GitHub.  For example, the
developers created the git repository for \textit{3D Slicer} in 2002, but did
not upload a copy of it to GitHub until 2020. Some GitHub data can be found
using its GitHub Application Program Interface (API) via the following url:
\textit{https://api.github.com/repos/[owner]/[repository]} where [owner] and
[repository] are replaced by the repo specific values. The number of months a
repository has been on GitHub helps us understand the average change of metrics
over time, like the average new stars per month.

@@

As in \citet{SmithEtAl2016}, Virtual machines (VMs) were used to provide an
optimal testing environment for each package. We used VMs because it is easier
to start with a fresh environment, without having to worry about existing
libraries and conflicts. Moreover, when the tests are complete the VM can be
deleted, without any impact on the host operating system. The most significant
advantage of using VMs is to level the playing field. Every software install
starts from a clean slate, which removes ``works-on-my-computer'' errors. When
filling in the measurement template, the grader notes the details for each VM,
including hypervisor and operating system version.

@@

The repository-based measurements summarize the information we can collect from
on-line resources. This information is incomplete because it doesn't generally
capture the development process, the developer pain points, the perceived
threats to software quality, and the developers' strategies to address these
threats.  Therefore, part of our methodology involves interviewing developers.

We based our interviews on a list of 20 questions, which can be found in
\citet{SmithEtAl2021}. Some questions are about the background of the software,
the development teams, the interviewees, and how they organize their projects.
We also ask about the developer's understanding of the users. Some questions
focus on the current and past difficulties, and the solutions the team has
found, or plan to try. We also discuss documentation, both with respect to how
it is currently done, and how it is perceived. A few questions are about
specific software qualities, such as maintainability, understandability,
usability, and reproducibility. The interviews are semi-structured based on the
question list; we ask follow-up questions when necessary.  The interview process
presented here was approved by the McMaster University Research Ethics Board
under the application number 
\href{https://github.com/smiths/AIMSS/blob/master/StateOfPractice/MACREM/Application.pdf}
{MREB\#: 5219}.

We sent interview requests to all 29 projects using contact information from
projects websites, code repository, publications, and from biographic pages at
the teams' institutions.  In the end nine developers from eight of the projects
agreed to participate: \textit{3D Slicer}, \textit{INVESALIUS 3}, \textit{dwv},
\textit{BioImage Suite Web}, \textit{ITK-SNAP}, \textit{MRIcroGL},
\textit{Weasis}, and \textit{OHIF}. We spent about 90 minutes for each
interview. One participant was too busy to have an interview, so they wrote down
their answers. In one case two developers from the same project agreed to be
interviewed. We held the meetings on-line using either Zoom or Teams, which
facilitated recording and automatic transcription. The full interview answers
can be found in \citet{Dong2021}.

@@

Table~\ref{tab_final_list} shows the 29 software packages that we measured,
along with summary data collected in the year 2020. We arrange the items in
descending order of LOC. We found the initial release dates (Rlsd) for most
projects and marked the two unknown dates with ``?''. The date of the last
update is the date of the latest update, at the time of measurement. We found
funding information (Fnd) for only eight projects.  For the Number Of
Contributors (NOC) we considered anyone who made at least one accepted commit as
a contributor. The NOC is not usually the same as the number of long-term
project members, since many projects received change requests and code from the
community.  With respect to the OS, 25 packages work on all three OSs: Windows
(W), macOS (M), and Linux (L). Although the usual approach to cross-platform
compatibility was to work natively on multiple OSes, five projects achieved
platform-independence via web applications. The full measurement data for all
packages is available in \citet{Dong2021-Data}.
